\documentclass[14pt, aspectratio=169]{beamer}
\usepackage[utf8]{inputenc}
\usetheme{default}
\useoutertheme{infolines}
\setbeamertemplate{navigation symbols}{}
% \usecolortheme{owl}
\setbeamertemplate{headline}{}

\title{Problems Faced in Programming for Research}
\subtitle{Causes, Effects and Some Strategies\\
\scriptsize (v0.1.0)}
\author{Cristian Dinu}
\date{March 2023}
\begin{document}

\frame{\titlepage}

\begin{frame}
\frametitle{Introduction}
\begin{itemize}
\item Scientific programming is still programming
\item The main issues of general programming still apply
    \begin{itemize}
    \item Unclear requirements
    \item Tough stakeholder management
    \item Bad planning
    \end{itemize}
\item But I will talk about problems specific to programming for research in my non-Large Hadron Collider experience
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Both research and coding are hard}
So that one individual can only do well one or the other
\begin{itemize}
\item Effects:
    \begin{itemize}
        \item It difficult to collect requirements and understand the problem domain
        \item (civilised) Friction and frustration between the researcher and the coder.
        \item Researchers stuck with old tools and methods (Excels with 100 sheets and colour codes)
    \end{itemize}
\item Strategies: "Embed" the developer in the team. Iterate fast, show and validate early, using friendly formats (Jupyter, Excel), avoid computer/programming slang, assume educator role
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Research projects are, well... research}
Requirements, goals, methods, change continuously as they progress. Ideas come up, real life is muddy, data is unstructured. 
\begin{itemize}
\item Effects: Nearly impossible to estimate and meet deadlines, frustration as a lot of time needs to be spent on menial task
\item Strategies: Prepare "plan B"-s, embrace changes, accept this is life, be happy with it, as it means progress.
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{The "Paper" is more important than the code}
The incentives and priorities are conflicting. Recognition not always given.
\begin{itemize}
\item Effects: Quick and dirty, duct-tape coding, neglected documentation, testing, maintainability, security and ethics reviews, and reproducibility
\item Strategies: Recognise software contributions, emphasise code quality, make code as important as the paper
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{People who code seem "in transit" to better roles or self-taught\ldots}
I saw most of the code is done "for free" by masters or PhD students or enthusiastic self-taught young academics.
\newline
\begin{itemize}
    \item Effects: Lack of good practices, inconsistent code quality, slow skills improvement. Lots of technical debt
    \item Strategies: Educate beyond coding -- e.g. source control, documentation, and establish some standards (e.g. everything goes on departments' GitHub account, no Excel allowed)
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\ldots because good developers are expensive and industry pays well}
Also, it's unfair to pay a developer three times as much as a researcher, just because the IT pays well.
\begin{itemize}
\item Effects: Difficulty attracting and retaining skilled engineers, reliance on inexperienced developers
\item Strategies: Allow people to be paid directly from grants, or projects with the industry. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Ethics, Security, and Privacy are a second thought}
In research programming, ensuring ethical conduct, maintaining security, and respecting privacy are crucial, but they are often overlooked or neglected.
\begin{itemize}
\item Effects:
\begin{itemize}
    \item Inadequate protection of sensitive data or private information
    \item Unintended consequences of research findings or technology use
    \item Misuse of research results or tools for malicious purposes
\end{itemize}
\item Strategies: educate (tell scary stories), create checklists so that important issues are not missed, check where code and data are stored and deployed.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{More and more blackboxes (libraries, AI models)}
\begin{itemize}
\item Effects: Reduced transparency, harder debugging, validation issues, over-reliance on external libraries
\item Strategies: Encourage open-source development, require explainable AI, prioritize validation and testing
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Programming for research shares difficult problems with "general" programming 
\item Competition with the industry is unfair [compared to academia]
\item Hopefully a "software is an investment not an expense" mindset will be adopted by the majority of people involved in research
\item But keep a WindowsXP machine handy. You might need it for next project.
\end{itemize}
\end{frame}

\end{document}
